---
title: "The BayesRMST R package"
author: "K Shuvo Bakar, and Armando Teixeira-Pinto"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
---

## Introduction

The `BayesRMST` package provides Bayesian methods for restricted mean survival time (RMST) analysis and adaptive selection of the restriction time in time-to-event studies. The package is designed for researchers and practitioners seeking clinically interpretable treatment effect estimates while accounting for uncertainty in survival modelling and restriction-time selection. The primary goals of the package are to:

* Provide Bayesian estimation of restricted mean survival time using a range of survival modelling approaches.
* Implement adaptive restriction-time selection based on posterior utility optimisation, rather than relying on arbitrary or ad hoc choices of the restriction time.
* Quantify uncertainty in both survival estimation and restriction-time selection within a unified Bayesian framework.
* Support incorporation of expert-elicited clinically meaningful time windows into the restriction-time optimisation procedure.
* Provide a consistent interface for Bayesian RMST analysis across multiple classes of survival models.

The package currently supports:

* Bayesian nonparametric survival models.
* Bayesian M-spline survival models.
* Bayesian parametric survival models.

These methods build upon established Bayesian and classical survival analysis approaches, with wrapper functions providing a unified interface to existing **R** implementations of nonparametric, M-spline, and parametric survival models. The package extends these methods by implementing a novel Bayesian optimisation framework for selecting the restriction time, enabling principled RMST inference that accounts for posterior uncertainty and clinically relevant decision criteria.

## Installation

### Prerequisites

Before installing `BayesRMST`, users must ensure that:

1. A working C++ toolchain/compiler is installed.
2. The `rstan` package is successfully installed and configured.

Since `BayesRMST` relies on Stan for Bayesian computation, proper compiler and `rstan` installation are required. For details see the README of `BayesRMST` here: https://github.com/ksbakar/BayesRMST

### Installation via GitHub

You can install the latest official release version via GitHub, which requires R pacakge `devtools` or `remotes`.

```{r install_github1, eval=FALSE}
# install.packages("devtools")
devtools::install.packages("ksbakar/BayesRMST")
```

or  

```{r install_github2, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ksbakar/BayesRMST")
```

The above commands download and install the latest version of the package from the GitHub repository. Once installed, the package can be loaded into the R

## Illustrative Examples


```{r setup}
library("BayesRMST")
```

### Main Function

The primary function implemented in the package is `survRMST()`, which performs Bayesian RMST analysis under several modelling frameworks. The function supports parametric, spline-based, and nonparametric Bayesian survival models, while also providing posterior inference and Bayesian decision-making tools.

The main arguments of the `survRMST()` function are summarised below.

- `data`: A data frame containing the survival data. The columns must be ordered as:
  1. Survival time.
  2. Event indicator/status coded as `0` or `1`.
  3. Treatment group coded as `0` or `1`.

- `modelType`: Specifies the Bayesian survival model to be fitted. Available options are:
  - `"BayesPara"` — Bayesian parametric model.
  - `"BayesMspline"` — Bayesian M-spline model.
  - `"BayesNonPara"` — Bayesian nonparametric model.

- `family`: Applicable only when `modelType = "BayesPara"`. The default distribution is Weibull, although any distribution supported by `brmsfamily` may also be specified.

- `iter`: Total number of Markov chain Monte Carlo (MCMC) iterations.
- `chains`: Number of MCMC chains to run.
- `cores`: Number of processor cores used for parallel computation.
- `control`: A named list of tuning parameters controlling the Stan sampler.
- `seed`: Random seed used to ensure reproducibility of results.
- `prior_expert_list`: A list specifying expert prior information on the restriction time parameter, $\tau$. For example,

```{r}
list(
  c(5, 6),
  c(10, 12),
  c(8.5, 10)
)
```
- `eta`: Decision threshold used in Bayesian power calculations. The default value is `0.95`.

The `survRMST()` function returns several outputs containing posterior summaries and Bayesian decision measures. The major outputs include:

- `rmst`: A table containing the estimated optimal RMST values together with posterior summaries.
- `model_fit`: The fitted Bayesian model object.
- `decision`: Bayesian power or decision outcome based on the specified threshold, $\eta$.

In addition to the main analysis function, the package also provides several S3 methods, including `print()` and `plot()`, to facilitate summarisation and visualisation of model outputs. These methods allow users to inspect posterior RMST estimates, credible intervals, treatment differences, and the optimal restriction time in a user-friendly manner.

### Datasets

The **BayesRMST** package can be applied to a wide range of survival datasets. In this vignette, we illustrate its use with three commonly used datasets available in existing **R** packages:

- `cetux` from **survextrap**, containing head and neck cancer survival data.
- `cancer` from **survival**, containing lung cancer survival data.
- `aml` from **survival**, containing acute myelogenous leukemia survival data.

For each dataset, the treatment variable is recoded to a binary indicator (`0`/`1`) before analysis.

```{r data-preparation}
# Cetuximab data
data(cetux, package = "survextrap")
cetux <- cetux[, c("years", "d", "treat")]
cetux$treat <- as.numeric(cetux$treat) - 1

# Lung cancer data
data(cancer, package = "survival")
cancer <- cancer[, c("time", "status", "sex")]
cancer$sex <- cancer$sex - 1
cancer$status <- cancer$status - 1

# Acute myelogenous leukemia data
data(aml, package = "survival")
aml$x <- abs(aml$x - 2)
```

## Bayesian M-spline Model

Fit the Bayesian M-spline model by specifying `modelType = "BayesMspline"`.

```{r mspline-fit}
out <- survRMST(
  data = cetux,
  modelType = "BayesMspline",
  prior_expert_list = list(c(3, 5), c(2, 3))
)

print(out)
```

The fitted object contains posterior RMST estimates together with the optimal restriction time under both the data-driven and elicited approaches.

Plot the estimated survival curves and RMST summaries using

```{r mspline-plot, fig.width=7, fig.height=5}
plot(out)
```

To display the signal-to-noise ratio used to select the restriction time, use

```{r mspline-omega, fig.width=7, fig.height=5}
plot(out, omega = TRUE)
```

## Bayesian Nonparametric Model

Fit the Bayesian nonparametric model by specifying `modelType = "BayesNonPara"`.

```{r nonparametric-fit}
out <- survRMST(
  data = cancer,
  modelType = "BayesNonPara",
  prior_expert_list = list(c(300, 500), c(800, 900))
)

print(out)
```

Visualise both the survival curves and the signal-to-noise ratio.

```{r nonparametric-plots, fig.width=10, fig.height=5}
p1 <- plot(out)
p2 <- plot(out, omega = TRUE)

gridExtra::grid.arrange(p1, p2)
```

## Bayesian Parametric Model

Fit a Bayesian parametric survival model by setting `modelType = "BayesPara"` and specifying a survival distribution.

```{r parametric-weibull}
out <- survRMST(
  data = aml,
  modelType = "BayesPara",
  family = "weibull",
  prior_expert_list = list(c(40, 60), c(60, 80))
)

print(out)
```

Alternative parametric distributions supported by **brms**, such as the exponential distribution, can also be used.

```{r parametric-exponential}
out <- survRMST(
  data = aml,
  modelType = "BayesPara",
  family = "exponential",
  prior_expert_list = list(c(40, 60), c(60, 80))
)

print(out)
```

These examples demonstrate the typical workflow in `BayesRMST`: fit a Bayesian survival model, estimate the optimal restriction time, summarise the results with `print()`, and visualise the fitted model using `plot()`.

## References

Bakar, K. S., Teixeira-Pinto, A., Tchetgen, E. J. T., Wong, G. (2026). *Data Adaptation and Elicitation for Restricted Mean Survival Time within a Bayesian Framework*. \emph{Submitted}.

